An Efficient Minibatch Acceptance Test for Metropolis-Hastings

### 2016-10-01 nips_2016.tex
Did about 1/3 of my desired changes to Appendix A

\subsection{A Worst-Case Gaussian Example}\label{ssec:gaussian_example}
Consider a Gaussian model where $x_1,\ldots,x_N \simiid \mathcal{N}(\theta,1)$
with a known variance $\sigma^2=1$, true mean $\theta=0.5$, and a uniform prior
on $\theta$. The log likelihood ratio is
\begin{equation}
\Lambda^*(\theta,\theta') =  \frac{N}{b}\sum_{i=1}^b \log\frac{p(x_i|\theta')}{p(x_i|\theta)}=
  %%\frac{1}{2}\left( (x_i-\theta')^2 - (x_i - \theta)^2\right) =
  N(\theta'-\theta)\left(\frac{1}{b}\sum_{i=1}^b x_i-\theta-\frac{\theta'-\theta}{2}\right)
\end{equation}
which is normally distributed over selection of the Normal samples $x_i$.  Since
the $x_i$ have unit variance, their mean has variance $1/b$, and the variance of
$\Lambda^*(\theta,\theta')$ is $\sigma^2(\Lambda^*) = (\theta'-\theta)^2N^2/b$.
In order to pass a hypothesis test that $\Lambda > \psi$, there needs to be a
large enough gap (several $\sigma(\Lambda^*)$) between
$\Lambda^*(\theta',\theta)$ and $\psi(u,\theta',\theta)$.

@@

Assume $(\theta' - \theta) \in \pm\frac{1}{\sqrt{N}}[0.5,1]$ and $\theta -0.5
\in \pm\frac{1}{\sqrt{N}}[0.5,1]$ with matching sign. These events occur with
probability ({\color{blue} Daniel: sorry for interjecting here but don't we need
to take into account the $1/\sqrt{N}$ in the probabilities? If $N \to \infty$
(and it's going to be large with big datasets) the probabilities should be about
zero.}) $p_0=2*(\Phi(1)-\Phi(0.5)) = 0.2997$ and
$p_1=(\Phi(1)-\Phi(0.5))=0.14988$ respectively, and guarantee that
$\Lambda^*(\theta',\theta)$ is negative. This then guarantees that we can find a
$u \in [0,1]$ so that $\psi(u,\theta',\theta)$ equals
$\mathbb{E}[\Lambda^*(\theta',\theta)]$.  Specifically, choose $u_0$ to satisfy
\begin{equation}\label{eq:log_uo}
    \log u_0 = N(\Lambda^*(\theta',\theta)-\psi(1,\theta',\theta))
\end{equation}
which evaluates ({\color{blue} Daniel: sorry again for interjecting but I really
think $\log u_0$ does not have that $N$ there. Also, for the subsequent
equation, I think we need to be using \emph{expected} values of $\Lambda^*$}) to
\begin{equation}
    \log u_0 = -N(\theta'-\theta)\left(\theta-0.5+\frac{\theta'-\theta}{2}\right).
\end{equation}
This means we can rewrite the acceptance test as
\begin{equation}
    \frac{1}{b}\sum_{i=1}^b x_i-\left(\theta+\frac{\theta'-\theta}{2} + \frac{\log u_0}{N(\theta'-\theta)}\right)
    \hspace{0.1in} \not\approx \hspace{0.1in} 0
\end{equation}
where $\not\approx$ means ``significantly different from'' under the
distribution over samples of $x_i$.  The above choice of $u_0$ ensures that the
terms in parenthesis above sum to 0.5. Since the $x_i$ have mean 0.5, the test
will never correctly succeed.  Furthermore, if we sample values of $u$ near
enough to $u_0$, the terms in parenthesis will not be sufficiently different
from 0.5 to allow the test to succeed.

>>>>

\subsection{A Worst-Case Gaussian Example}\label{ssec:gaussian_example}
Consider a Gaussian model where $x_1,\ldots,x_N \simiid \mathcal{N}(\theta,1)$
with a known variance $\sigma^2=1$, true mean $\theta=0.5$, and a uniform prior
on $\theta$. The log likelihood ratio is
\begin{equation}
\Lambda^*(\theta,\theta') =  \frac{N}{b}\sum_{i=1}^b \log\frac{p(x_i|\theta')}{p(x_i|\theta)}=
  N(\theta'-\theta)\left(\frac{1}{b}\sum_{i=1}^b x_i-\theta-\frac{\theta'-\theta}{2}\right)
\end{equation}
which is normally distributed over selection of the Normal samples $x_i$.  Since
the $x_i$ have unit variance, their mean has variance $1/b$, and the variance of
$\Lambda^*(\theta,\theta')$ is $\sigma^2(\Lambda^*) = (\theta'-\theta)^2N^2/b$.
In order to pass a hypothesis test that $\Lambda > \psi$, there needs to be a
large enough gap (several $\sigma(\Lambda^*)$) between
$\Lambda^*(\theta',\theta)$ and $\psi(u,\theta',\theta)$.

@@

{\color{blue} Daniel: changes I made:

\begin{enumerate}
    \item Changed all occurrences of $\Lambda^*(\theta',\theta)$ to
    $\Lambda^*(\theta,\theta')$, i.e., switched the input ordering.

    \item Originally, $\log u_0$ was defined as
    $N(\Lambda^*(\theta,\theta')-\psi(1,\theta,\theta'))$ but I changed the
    starting definition so it does \emph{not} have a factor of $N$ in its
    definition (because we later get $N$ to appear in a subsequent step).

    \item In addition, $\psi(1,\theta,\theta') = 0$ due to uniform priors and
    symmetric proposals and there didn't seem to be a good enough reason to
    include it in the definition so I deleted it (sometimes it's insightful to
    have a nice form of zero inside, but I didn't see it here).

    \item I added in an ``$E$'' because it seemed like you wanted the ``$u$''
    that would make $\psi(u,\theta,\theta') = \log u =
    E[\Lambda^*(\theta,\theta')]$.
\end{enumerate}

I'm hoping this is correct. I understood the general idea of what you were
trying to do, it's a matter of getting the painful interval arithmetic details
down.
}

Assume $(\theta' - \theta) \in \pm\frac{1}{\sqrt{N}}[0.5,1]$ and $\theta -0.5
\in \pm\frac{1}{\sqrt{N}}[0.5,1]$ with matching sign. These events occur with
probability ({\color{blue} Daniel: sorry for interjecting here but don't we need
to take into account the $1/\sqrt{N}$ in the probabilities? If $N \to \infty$
(and it's going to be large with big datasets) the probabilities should be about
zero.}) $p_0=2*(\Phi(1)-\Phi(0.5)) = 0.2997$ and
$p_1=(\Phi(1)-\Phi(0.5))=0.14988$ respectively, and guarantee that
$\Lambda^*(\theta,\theta')$ is negative. This then guarantees that we can find a
$u \in [0,1]$ so that $\psi(u,\theta,\theta') = \log u$ equals
$E[\Lambda^*(\theta,\theta')]$.

Specifically, choose $u_0$ to satisfy $\log u_0 = E[\Lambda^*(\theta,\theta')]$.
Using $E[x_i] = 0.5$, we see that
\begin{align}
    \log u_0 &= N(\theta'-\theta)\frac{1}{b} \cdot E\left[\sum_{i=1}^b x_i-\theta-\frac{\theta'-\theta}{2}\right] \\
    &= -N(\theta'-\theta)\left(\theta-0.5+\frac{\theta'-\theta}{2}\right).
\end{align}
This means we can rewrite the acceptance test as
\begin{equation}
    \frac{1}{b}\sum_{i=1}^b x_i-\left(\theta+\frac{\theta'-\theta}{2} + \frac{\log u_0}{N(\theta'-\theta)}\right)
    \hspace{0.1in} \not\approx \hspace{0.1in} 0
\end{equation}
where $\not\approx$ means ``significantly different from'' under the
distribution over samples of $x_i$.  The above choice of $u_0$ ensures that the
terms in parenthesis above sum to 0.5. Since the $x_i$ have mean 0.5, the test
will never correctly succeed.  Furthermore, if we sample values of $u$ near
enough to $u_0$, the terms in parenthesis will not be sufficiently different
from 0.5 to allow the test to succeed.
